A Framework for Discrete Change

### 2008-11-26 papers/individual/individual.tex
updated individual paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@244 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{A class of models is developed for measuring and detecting
discrete change in learning and development.  The basic model for
detecting such change is the latent or hidden Markov model.
Traditionally, these models were restricted to categorical, mostly
binary, observed variables, placing severe restrictions on possible
measurement models.  In this paper, the basic model is extended to
include arbitrary distributions for the observed variables, including
multi-variate distributions.  Moreover, there is optional support to
include time-varying predictors.  In effect, this model consists of
mixtures of generalized linear models with Markovian dependencies over time
to model the change process.  In addition, transition parameters can
be made to depend on covariates as well, such that the switching
regime between states depends on characteristics of the individual or
the experimental situation.  The model is illustrated with an example
of participants' learning in the Iowa Gambling Task and in the Weather 
Prediction Task.}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts, in performance on Piagetian tasks, in
discrimination learning and in conditioning.  This chapter is
concerned with detecting the nature and time points of change in (individual)
time series.  We present a framework of dependent mixture models that can
be used to differentiate between gradual and discrete learning events
in individual time series data.  Before presenting the model in formal
terms and providing some illustrations, we first review some examples
in which discrete change is found.

% phenomena of discrete change in development/learning
Piagetian developmental theory assumes step-wise changes in the
strategies that children apply in all kinds of tasks such as the
conservation learning and the balance scale task (REFERENCE??).
\citet{Maas1992} developed a catastrophe model to describe step-wise
changes or phase transitions in learning and developmental processes.
They applied the catastrophe model to learning in the conservation of
liquid task (REFERENCE??)  in which children have to judge relative
volumes of liquid in glasses of different heights and widths.  Young
children tend to ignore the width dimension and hence always choose
the glass with the highest level of liquid (REFERENCE??).
\citet{Maas1992} showed that there is a sudden transition to a new
strategy in which the children also take the width of the glasses into
account when judging the volume of liquids.

% balance scale rules
\citet{Jansen2001} applied the catastrophe model to
development of strategies on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

% hysteresis on the balance scale
\citet{Jansen2001} found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, she found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  \citet{Jansen2001} also found
evidence for hysteresis: the phenomenon that switching between
strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

>>>>

\abstract{A class of models is developed for measuring and detecting
discrete change in learning and development.  The basic model for
detecting such change is the latent or hidden Markov model.
Traditionally, these models were restricted to categorical, mostly
binary, observed variables, placing severe restrictions on possible
measurement models.  In this paper, the basic model is extended to
include various distributions for the observed variables, including
mixed multi-variate distributions.  Moreover, there is an option to
include time-varying predictors on the observed distributions.  In
effect, this model consists of mixtures of generalized linear models
with Markovian dependencies over time to model the change process.  In
addition, the transition parameters can be estimated with covariates,
such that the switching regime between states depends on
characteristics of the individual or the experimental situation.  The
model is illustrated with examples of participants' learning in the
Iowa Gambling Task and in the Weather Prediction Task.}

@@

% intro
Discrete change frequently occurs in learning and development: in
learning concepts and discrimination learning, in performance on
Piagetian tasks such as the conservation of liquid task and the
balance scale task, and in conditioning.  This chapter is concerned
with detecting and characterizing such discrete changes in
(individual) time series data.  We present a framework of dependent
mixture models that can be used to test whether discrete learning
events are present and, if so, to test at which point in time those
changes are taking place.  Before presenting the model in formal terms
and providing some illustrations, we first review some examples from
the developmental psychology literature in which discrete change is
typically found.

% phenomena of discrete change in development/learning
Piagetian developmental theory assumes step-wise changes in the rules
that children apply in all kinds of tasks such as in conservation
learning and in the balance scale task (SieglerAlibali19??).
\citet{Maas1992} developed a catastrophe model to describe step-wise
changes predicted to occur by Piagetian theory; importantly, they
derived precise statistical criteria for testing whether two
discretely different stages exist.  They applied the catastrophe model
to development on the conservation of liquid task (InhelderPiaget19??)
in which children have to judge relative volumes of liquid in glasses
of different heights and widths.  Young children tend to ignore the
width dimension and hence always choose the glass with the highest
level of liquid (InhelderPiaget19??).  It is assumed that there is a
sudden transition to a new rule, for example, a rule in which the
children also take the width of the glasses into account when judging
the volume of liquids.

% \citet{Maas1992} 
% opzoeken: Siegler & Alibali 19??, intro textbook
% opzoeken: Inhelder & Piaget, 19??

% balance scale rules
\citet{Jansen2001} applied the catastrophe model to
development of reasoning on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

% Jansen 2001: aanpassen in de paper ref ipv het proefschrift

% hysteresis on the balance scale
\citet{Jansen2001} found evidence for stage-wise transitions, or
discrete change in our model, between Rule 1 and Rule 2 by testing
criteria that were derived from the catastrophe model.  In particular,
she found bimodal test scores and inaccessibility.  The latter means
that there are no in-between strategies: children apply either Rule 1
or Rule 2 and there is no in-between option.  \citet{Jansen2001} also
found evidence for hysteresis: the phenomenon that switching between
strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different values of a continuously
changing independent variable.  In particular, if the distance
dimension in the balance scale problems is made increasingly more
salient by increasing the distance difference between weights on
either side of the balance scale, children may switch from Rule 1 to
Rule 2.  If subsequently the distance difference is decreased again,
children may switch back to using Rule 1.  Hysteresis is the
phenomenon that this switch back occurs at a different value of the
control variable, in this case the distance difference.
